### 2020-03-04 doc/icfp20/counterpoint.tex
tweak comparison to Haskell '17 work

The type-theoretical formalization of counterpoint has first
been attempted by \citet{szamozvancev-welltyped}.
They encode the rules for intervals and motion by declaring
type classes \texttt{ValidHarmInterval} and \texttt{ValidMotion},
and specifying which instances are invalid using GHC's support
for custom type errors.

The availability of the custom error feature is an advantage of using
Haskell, but the language's type system does not allow producing
detailed error messages.
Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
an octave (8th) of D, breaking the rule that perfect intervals cannot
be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent Observe that the message clearly tells us which intervals
form what kind of invalid motion.

In Szamozvancev and Gale's implementation, the same example
results in the following error:

\begin{alltt}
Direct motion into a perfect octave is forbidden.
\end{alltt}

\noindent The lack of the precise information about the intervals
and motion is due to the absence of native dependent types in
Haskell.

Our representation of counterpoint improves on their
implementation in two ways.
First, while \citet{cong-cp} rely on the Agda type checker to report
errors in the counterpoint, we use our own type checker (implemented
as the \texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce helpful error messages while keeping
types simple.
Indeed, for the parallel octave example discussed above,
\citet{cong-cp} would just report the existence of an unsolved
metavariable \texttt{\_13}, meaning that Agda failed to construct
a proof required by the counterpoint constructor:

@@

The second improvement from \citet{cong-cp} is that, instead of
incorporating all the rules into a counterpoint datatype, we define
each of them as a separate function and combine them using a
record type.
This makes it easy to switch between strict and relaxed rule sets:
we just need to add or remove certain fields.
It also allows us to reuse some of the rules in a context other than
counterpoint, as we will see in the next section.

>>>>

The type-theoretical formalization of counterpoint has first been
attempted by \citet{szamozvancev-welltyped}.
They encode the rule for intervals by declaring a type class
\texttt{ValidHarmInterval} that classifies valid intervals, and then
specifying which instances are invalid using GHC's support for
custom type errors.
They also implement the restriction on motion based on a similar idea.

The availability of the custom type error feature is an advantage of
using Haskell: it allows the rule writer to leverage Haskell's type checker.
We believe that extending Agda with such support would be
beneficial in music formalization and for many other purposes.

Our representation of counterpoint improves on Cong and Leo's
in two ways.
First, while Cong and Leo rely on the Agda type checker to report
errors, we use our own type checker (implemented as the
\texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce more user-friendly error messages.
Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
an octave (8th) of D, breaking the rule that perfect intervals cannot
be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent In contrast, Cong and Leo would just report the existence
of an unsolved metavariable \texttt{\_13}, meaning that Agda failed
to construct a proof required by the counterpoint constructor.
Thus, the user would see the following message in the Agda buffer:

@@

The second improvement from \citet{cong-cp} is that, instead of
incorporating all the rules into a counterpoint datatype, we define
each rule as a separate function and combining them using a record type.
This makes it easy to switch between strict and relaxed rule sets:
we just need to add or remove certain fields.
It also allows us to reuse some of the rules in a context other than
counterpoint, as we will see in the next section.
